Dynamics of Anthropomorphism in Human-Robot Interaction

### 2014-11-24 paper.tex
In depth pass on the second part (section 3 and 4)

@@

\newcommand{\antNorm}[1][]{%
      \ifthenelse{\isempty{#1}}%
      {\widehat{\mathcal{A}}}
      {\widehat{\mathcal{A}}(#1)}
}

\newcommand{\ICA}{{$\mathcal{A}_0$~}}
\newcommand{\SLA}{{$\mathcal{A}_\infty$~}}
\newcommand{\sla}{{\mathcal{A}_\infty}}
\newcommand{\AntMax}{{$\mathcal{A}_{max}$~}}
\newcommand{\antMax}{{\mathcal{A}_{max}}}

@@

So far, the HRI community has not much investigated how anthropomorphism in
human-robot interactions evolves over time (during the process of
\emph{adopting} a robot, for instance). Though conflicts were found in the
understanding of what anthropomorphism is, the HRI community more or less
accepted viewing anthropomorphism as a static feature that once observed during
a short-term interaction reflects a sustaining social effect. However, if we
agree that anthropomorphism arises in an interaction, and thus reflects a
specific relation, we cannot accept it as a static phenomenon.

Similarly, \citet{kanda_interactive_2004} point out that relationships among
people evolve over time and, in a similar manner, one can expect people's
attitude toward technological artifacts and their relationship with them to
evolve as well.  In their study, \citet{kanda_interactive_2004} were among the
first to describe empirically \emph{novelty effects} during a long-term study
with an interactive robot in a school.

From a theoretical perspective, we also expect anthropomorphism to change over
time, with growing experience with a robot. As discussed in the previous
sections, anthropomorphism means in essence \emph{reasoning about and perceiving
    something non-human (and unknown or unfamiliar) based on one's
representation of the familiar and well-known concept of being human (or
one-self)}. According to \citet{epley_when_2008}, the basic operations that
underlie these inductive inferences are the \emph{acquisition of knowledge}, the
\emph{activation or elicitation of knowledge}, and the \emph{application of
activated knowledge at the time of the judgment}. By applying knowledge to a
given situation or interaction, one also attempts to correct, adjust, or maybe
integrate less accessible information into a more automatically activated
default representation.  This process of refining the applied knowledge are
natural because a person's knowledge base changes constantly with newly acquired
things. For instance, a person can acquire new knowledge about a robot when
interacting with it. With continued interaction with the unknown system also the
user's expertise with the system grows, and it can be expected that in turn, the
user will become able to better predict the previously unknown system.
Consequently, as the robot becomes more predictable, there is less uncertainty,
and the ``need to anthropomorphize'' the robot decreases (from the
human-centered point of view).

In other words, humans have a motivation to interact effectively in their
environment \textit{(effectance)} \citep{epley_when_2008}.  This means that
humans are motivated to be able to understand and predict the environment and
the agents around them. Thus, a person will seek to become familiar with what is
unfamiliar. Anthropomorphism is encouraged when an agent is unfamiliar.
Consequently, in the process of familiarizing oneself with an agent, the
tendency to anthropomorphize this agent should decrease.

We can not be sure about how exactly the process of familiarization to an
unfamiliar agent (based on continued interaction with the robot) looks like. In
turn, we can not make a precise statement about how continued interaction
decreases anthropomorphism but we suspect there is a link between the two.

In the introduction, we reported on a few selected experimental results that
illustrate how the user tendency to anthropomorphize decreased over time.
Combined with the extensive discussion on the nature and factors of
anthropomorphism that we proposed in section~\ref{sec:anthropomorphism}, we
introduce in the following section a novel qualitative model of the evolution of
anthropomorphism over time, \ie a model of the \emph{dynamics of
anthropomorphism}.

@@

The phenomenological model of anthropomorphism we propose
(Figure~\ref{fig:dynamics},~\citep{lemaignan2014dynamics}) represents how the
level of ``anthropomorphic effects''~\Ant[R,H,C,t] (\ie the observable
manifestations of anthropomorphism) evolves over a long-term human-robot
interaction.\footnote{We cannot make a precise statement about how long
``long-term'' interaction has to be to be considered as such. It seems to be
practical to not speak of ``several days'' or ``weeks'' but to rather consider
the duration and frequency of the actual interactions with the artifact. One
could probably call this the ``interaction history''.}

While the model primarily considers \emph{long-term interaction} -- direct
(non-mediated), repeated interaction with the same robot, over an extended
period of time (typically longer than a week), we also formally introduce a
so-called \emph{novelty effect}~\citep{kanda_interactive_2004} that models the
first phase of human-robot interaction, during which a specific increase of
anthropomorphic interactions is observed.

In the \emph{Dynamics of Anthropomorphism} model, anthropomorphism is quantified
by a \emph{normalized level of anthropomorphic effects} $\antNorm[R,H,C,t] =
\frac{\ant[R,H,C,t]}{\antMax}$: because anthropomorphic effects are often
qualitative and difficult to quantify on an absolute scale, we present them as a
normalized value, that spans from a minimum (no anthropomorphic effects:
$\ant=0$) to a maximum (noted \AntMax, with $t_{max} = \operatorname{arg\,max}_t
\, \ant[t]$). For the sake of readability, we will use \Ant in the following
sections to refer to the normalized level of anthropomorphic effects. Note also
that we discuss in section~\ref{sec:measuring} possible methodologies for the
quantitative and qualitative assessment of anthropomorphic effects.

The actual maximum value of anthropomorphic effects depends on each unique
combination of a human, a robot, a situation and several other factors we
introduce below. Thus, the maximum value varies according to the specific
characteristics of these factors. Figure~\ref{fig:dynamics} pictures this
maximum value at $t_{max}$, which actually corresponds to what we call the
novelty effect peak.\footnote{We assume that there is a peak in the human's
experiences of the novelty effects of the unfamiliar system. Further, we suspect
that this novelty peak comes along with the maximum value of anthropomorphic
effects.} The general \emph{shape} of the curve remains however the same and
depicts the evolution of anthropomorphism over time, \ie the general dynamics of
anthropomorphism.

@@

\draw (-0.05, 3) -- (0.05, 3) node[anchor=east] {\ICA};
\draw (-0.05, 5) -- (0.05, 5) node[anchor=east] {\AntMax};

% vertical axis - end
\draw[->] (9.8,0) -- (9.8,4) node[anchor=east] {};
\draw (9.8, 0.8) node[anchor=west] {\SLA};

@@

\caption{\textbf{Conceptual model of the dynamics of anthropomorphism}. We
    distinguish three main phases: \emph{initialization}, \emph{familiarization}
    and \emph{stabilization}, preceded by a \emph{pre-interaction} phase. The
    pre-interaction phase is characterized through an \emph{initial capital of
    anthropomorphism} (ICA, noted \ICA).  Once the interaction starts, the level
    of anthropomorphism increases due to the \emph{novelty effect}, and then
    decreases to reach a \emph{stabilized level of anthropomorphism} (SLA, noted
    \SLA). \ICA and \SLA may vary depending on the user (and his/her
    expectations), the robot and the context of interaction.  During the
    interaction, unpredicted behaviors of the robot (\emph{disruptive
    behaviors}) may lead to local increases or decreases of the level of
    anthropomorphic effects. The duration of these phases may vary and the
    overall shape of the curve is unique to each human interacting with a
    specific robot in a given context, as depicted by the alternative curves.}

@@

\subsubsection{Initialization Phase\\}

The \textit{Initialization Phase} starts with the so-called \textit{Initial
Capital of Anthropomorphism} ($ICA$), which corresponds to \ICA (value of \Ant
at time point $0$). This value considers the multiple factors presented in
section~\ref{sec:multiple-factors}, thus it is composed by factors of the human
$H$, the robot $R$, and the context of use $C$. As such, \ICA describes the
value of anthropomorphic effects that can be expected to occur during the first
contact between the human and the robot in a given situation. During this first
phase, pre-conceptions and initial expectations as well as previous experience
of the user with other similar types of robots and similar contexts can have a
strong impact and influence the general tendency to anthropomorphize.

During the \textit{Initialization Phase} we expect an increasing value of $A$
that may be partly due to the so-called \textit{effects of novelty} of the
robot. Such effects have been described with interactive technologies in general
\citep{rogers_diffusion_1995}, and also in several long-term HRI studies in
workplaces \citep{huttenrauch_fetch-and-carry_2003,mutlu_robots_2008}, in
schools and public places
\citep{gockley_designing_2005,kanda_communication_2005,kanda_interactive_2004},
and elder care centers \citep{sabelli_conversational_2011}. Also long-term
studies of domestic robots have shown similar effects and found that after some
time people's interest, engagement, and fascination of the robot decreased
\citep{sung_robots_2009,sung_domestic_2010,fernaeus_how_2010,fink_living_2013}.

What type of anthropomorphic effects are typical during the
\textit{Initialization Phase}? In our own work, we observed that users
spontaneously tend to talk directly to the robot (though the user knows that the
system is not be able to recognize speech), for instance, by greeting the robot,
asking questions to it, giving it nick-names or ascribing a gender, praising the
robot, commenting on its performance in general, and giving commands. Also, the
use of point gestures and waving at the robot were fairly common among
participants of our studies, as well as ascribing gender and some personal
preferences or intention to the robot. (Note that we did not even use an
anthropomorphic robot but a Roomba vacuum cleaning robot.)

In general, it appears that anthropomorphism during the \textit{Initialization
Phase} is a commonly observed phenomenon. But we would like to stress again that
most of these effects are likely to fade out over time.

@@

In the \textit{Familiarization Phase} the user gets acquainted and familiar with
the robot and with using the robot, and hence the tendency to anthropomorpize
decreases. This phase typically starts as the local maximum value of
anthropomorphic effects that are enhanced by the novelty of the robot start to
wear off. The user may think that he/she faces an agent that aims at
``human-level intelligence'' (to take McCarthy's words). This is a transient
state that quickly vanishes as the user recognized that the behavior of the
robot is generally \textit{predictable} and possibly \textit{deceptive}. For
instance, the robot produces verbal utterances (it ``talks'') but it does not
understand when we talk; or the robot is endowed with eyes but does not
recognize the objects we show it.

@@

We call \emph{disruptive behaviors} any behavior exhibited by the robot hat
appears unexpected to the user: for instance, a robot may usually follow always
the same route to go from one place in a house to the other, but suddenly it
might change the route. The actual underlying reason may span from a bug to the
detection of a new obstacle, but as long as this reason is not immediately
intelligible to the user, the behavior counts as \emph{disruptive}. We further
discuss the role of disruptive behaviors in cognitive terms in section
\ref{sec:disruptive}.

@@

During the \textit{Stabilization Phase} the value of $A$ tends to stabilize over
a longer time, to reach a value that we call the \textit{Stabilized Level of
Anthropomorphism} ($SLA$), which corresponds to \SLA. The value that the
$SLA$ approaches is unique for each user interacting with a specific robot in a
given situation. \SLA may approach $0$, when no anthropomorphic effects
are observed anymore, for instance, when the user is interacting with the robot
in a routine way and he/she does not project any human-like traits unto the
robot\footnote{The robot may have become \textit{``invisible in use''}
\citep{takayama_toward_2011}.} As such, the $SLA$ stands for the long-term
lasting value of anthropomorphism.  We proposed that \SLA (the value
corresponding to the $ICA$) is built on multiple factors (see
section~\ref{sec:multiple-factors}) and similar \SLA is based on these
factors, however taking into account the \textit{interaction history} $t$ that
the user has with the robot.  Note that the $ICA$ and the $SLA$ are generally
not correlated. A person may initially have a high potential of
anthropomorphizing (high \ICA), for instance a highly anthropomorphic robot,
but subsequently get disappointed by the actual capabilities of it and stop
explaining it in terms of human-like characteristics (low \SLA). Another
user may however create lasting affective bonds with the same robot and may
continue to anthropomorphize the robot (high \SLA).

@@

\path[fill=gray!20] (0,0) rectangle (10.5,2);
\path[fill=gray!50] (0,2) rectangle (10.5,4);
\path[fill=gray!20] (0,4) rectangle (10.5,6);

\draw (-0.3,1) node[rotate=90] {Stage 1}
      (-0.3,3) node[rotate=90] {Stage 2}
      (-0.3,5) node[rotate=90] {Stage 3};

@@

\draw (11,1) node[align=left, anchor=west]{\scriptsize pre-cognitive\\\scriptsize anthropomorphism}; %label

\draw (11,3) node[align=left, anchor=west] {\scriptsize{projection of existing}\\\scriptsize{mental models}\\\scriptsize{\it (familiarity)}}; %label

\draw (11,5) node[align=left, anchor=west] {\scriptsize{adapted mental
model} \\ $\to$ \scriptsize{adapted interaction}\\\scriptsize{\it (anthropomorphic or not)}}; %label

@@

\end{tikzpicture}
%}
\caption{\textbf{The three cognitive stages of anthropomorphism}: Stage 1 is the instinctive,
sub-cognitive identification of living peers. {\it Empathy} is characteristic
of this stage. After longer observation or short, non-contextualized interaction
(typically, a lab environment), the user reaches Stage 2: the user projects a
mental model he/she is already familiar with onto the robot. After longer {\it
contextualized} interaction (typically, at home), the user enters Stage 3 of
anthropomorphism: the user composes a custom mental model of the robot,
based on experience. This leads to adapted interaction modalities, that may
still be anthropomorphic, or not.}
\label{fig:cognitivemodel}
\end{figure}

We propose three different cognitive stages (Figure~\ref{fig:cognitivemodel}),
which are not mutually exclusive but overlap over time. Note that these three
cognitive stages are related but do not exactly match the \emph{Initialization},
\emph{Familiarization} and \emph{Stabilization} phases introduced in our model
of the dynamics of anthropomorphism.

@@

The intuitive anthropomorphization in the cognitive Stage 1 might be
characteristic of what \cite{takayama_perspectives_2012} calls the
\textit{in-the-moment} perspective on agency with a robot compared to the
\textit{reflective} perspective (see paragraph ``artifact-centered perspective''
in section \ref{sec:anthropomorphism}).  For instance, we reflectively might
\textit{not} perceive agency in a social robot, but it can feel quite
differently in the moment of interaction. \cite{takayama_perspectives_2012}
argues that neglecting to separate reflective perspectives from in-the-moment
perspectives of agency is one of the major sources of confusion when people talk
and write about anthropomorphism. There seems to be a disconnection between what
people consciously perceive and how they respond to stimuli that they may not
consciously perceive. That means, in the initial phase of interacting with a
novel device \textit{(initialization)} (see Figure \ref{fig:dynamics}) people
might not respond consciously but rather mindlessly \citep{nass_machines_2000},
and in turn anthropomorphize more.  Only after some time of
\textit{familiarization}, they might respond in a more reflective manner (in
turn, the tendency to anthropomorphize might decrease). This can be illustrated
by the fact that people tend to deny interacting with computational systems as
if they were people and yet they respond to computers in many ways that are
remarkably similar to how they respond to people \citep{reeves_media_1996}.
\cite{takayama_perspectives_2012} also applies a cognitive viewpoint on the two
different perspectives of in-the-moment \vs reflective to illustrate
differences. She states that in-the-moment perceptions of agency are largely
shaped by bottom-up perceptual processes, evoking very immediate responses (\eg
to the cues emitted by the artifact's design (``artifact-centered
perspective''). In contrast, according to Takayama, reflective perceptions are
more often shaped by top-down processes because of the nature of reflective
thought.

After a longer observation period (typically including complete action sequences
of the robot) or short interaction (touching, short talk like greetings), we
suggest the human reaches the cognitive \emph{Stage 2}: in this phase, the human
starts building a behavioral and cognitive model of the robot that would support
both the observed and imagined capabilities of the robot.  The \emph{familiarity
thesis}~\cite{hegel_understanding_2008} supports the idea that the human first
projects onto the robot mental models of similar agents he/she is already
familiar with (ranging from animals to human adults, to pets and children). We
hypothesize that the nature of the projected mental model, as well as how deep
the human engages in this projection, might be driven by the same parameters as
we mentioned for the \emph{initial capital of anthropomorphism}
(section~\ref{sec:ica}).  At this stage the human might also change his attitude
towards the robot, paying more attention to social cue than to low-level
behavior which may reinforce the resonance mechanisms.

@@

This mental process depends on the human understanding of the robot's inner
working, as well as his/her own tendency to anthropomorphize (the
\emph{personality} in ICA factor), but at this stage, the \emph{perception} of
the robot (its shape for instance) and its intended \emph{purpose} play a less
important role. It is mostly a human-centric process.  The result of this third
phase would be an iteratively adapted cognitive model of the robot. It is also
at this last stage that the human might have a modified tendency to
anthropomorphize, because he/she has built a full cognitive model of the robot.
Consequently, the human would now be able to attribute appropriate internal
states to the robot through his/her theory of mind.

@@

As mentioned before, the three cognitive stages proposed here do not match with
the three phases in the \emph{Dynamics of Anthropomorphism}.  In particular, the
cognitive Stages 1 and 2 are both included in the \emph{initialization} phase of
the anthropomorphism model. Sub-cognitive anthropomorphism typically
\emph{initiates} the novelty effect by rapidly engaging the human in the
interaction through an initial projected agency, whereas cognitive Stage 2
(projection of familiar mental models) supports the novelty effect by inducing
beliefs that the robot is set up with possibly complex cognitive abilities.

>>>>

\newcommand{\AntE}[1][]{%
      \ifthenelse{\isempty{#1}}%
      {$\mathcal{A}^{e}$}
      {$\mathcal{A}^{e}(#1)$}
}

@@

\newcommand{\antE}[1][]{%
      \ifthenelse{\isempty{#1}}%
      {\mathcal{A}^{e}}
      {\mathcal{A}^{e}(#1)}
}

\newcommand{\antENorm}[1][]{%
      \ifthenelse{\isempty{#1}}%
      {\widehat{\mathcal{A}^{e}}}
      {\widehat{\mathcal{A}^{e}}(#1)}
}

\newcommand{\ICA}{{$\mathcal{A}_0$~}}
\newcommand{\ICAe}{{$\mathcal{A}^{e}_0$~}}
\newcommand{\SLA}{{$\mathcal{A}_\infty$~}}
\newcommand{\SLAe}{{$\mathcal{A}^{e}_\infty$~}}
\newcommand{\sla}{{\mathcal{A}_\infty}}
\newcommand{\AntMax}{{$\mathcal{A}_{max}$~}}
\newcommand{\AntEMax}{{$\mathcal{A}^{e}_{max}$~}}
\newcommand{\antEMax}{{\mathcal{A}^{e}_{max}}}

@@

So far, the HRI community has not investigated how anthropomorphism in
interactions with robots evolve over time. There are several reasons for why it
makes sense to consider temporal dynamics in anthropomorphism.

One reason for why changes in anthropomorphism over time are likely, comes from
the cognitive perspective, as outlined in the previous section. We are
spontaneous but also reflect on our interactions with the robot.  From the
artifact-centered perspective on anthropomorphism (thus, looking at the
characteristics of the robot) it is assumed that humans directly (more or less
mindlessly) respond to life-like and social cues that a non-human agent (here:
robot) emits~\citep{reeves_media_1996,nass_machines_2000}. Without huge
thoughtful mental processing, humans tend to apply stereotypes and heuristics to
the robot, and in turn apply human-human social schemas and norms to the
occurring interactions. However, this may only be some \textit{immediate}
(sometimes visceral) experience, reflecting the first impression that the user
has of the robot, based on its appearance~\citep{norman_emotional_2003}.
\cite{takayama_perspectives_2012} refers to this as the \textit{in-the-moment}
perspective we have on the robot, compared to a more distanced cogitation and
consideration, which the author calls \textit{reflective}.  Further, we also
become familiar with the robot as we continue interacting and dealing with it,
we gain knowledge about it, and refine our mental model of the robot.

Another reason for considering temporal dynamics are the changes in user
experience~\citep{karapanos_user_2009} and relationship to the robot.
\cite{kanda_interactive_2004} pointed out that not only the relationships among
people evolve over time but that, in a similar manner, one can expect that also
people's relationship and their attitude toward robots evolve. The authors were
among the first ones to describe empirically \textit{novelty effects} during a
long-term study with an interactive robot in a school.

In the introduction, we reported on a few selected experimental results that
illustrate how the user tendency to anthropomorphize decreased over time.
Combined with the discussion on the nature and factors of
anthropomorphism that we proposed in section~\ref{sec:anthropomorphism}, we
introduce in the following section a novel qualitative model of the evolution of
anthropomorphism over time, \ie a model of the \emph{dynamics of
anthropomorphism}.

@@

The phenomenological model of anthropomorphism we propose
(Figure~\ref{fig:dynamics},~\citep{lemaignan2014dynamics}) represents how the
level of anthropomorphic \emph{effects}~\AntE[R,H,C,t] (\ie the observable
manifestations of the anthropomorphism \Ant) evolves over a long-term human-robot
interaction.

While the model primarily considers \emph{long-term interaction} -- direct
(non-mediated), repeated interaction with the same robot, over an extended
period of time (typically longer than a week) --, we also formally introduce a
so-called \emph{novelty effect}~\citep{kanda_interactive_2004} that models the
first phase of human-robot interaction, during which a specific increase of
anthropomorphic interactions is observed.

In this model, anthropomorphism is quantified
by a \emph{normalized level of anthropomorphic effects} $\antENorm[R,H,C,t] =
\frac{\antE[R,H,C,t]}{\antEMax}$: because anthropomorphic effects are often
qualitative and difficult to quantify on an absolute scale, we present them as a
normalized value, that spans from a minimum (no anthropomorphic effects:
$\antE=0$) to a maximum (noted \AntEMax, with $t_{max} = \operatorname{arg\,max}_t
\, \antE[t]$). For the sake of readability, we will use \Ant in the following
sections to refer to the normalized level of anthropomorphic effects. Note also
that we discuss in section~\ref{sec:measuring} possible methodologies for the
quantitative and qualitative assessment of anthropomorphic effects.

The actual maximum value of anthropomorphic effects depends on each unique
combination of a human, a robot, a situation and several other factors we
introduce below. Figure~\ref{fig:dynamics} pictures this
maximum value at $t_{max}$, which we hypothesize to correspond to the
novelty effect ``peak''.

@@

\draw (-0.05, 2.1) -- (0.05, 2.1) node[anchor=east] {\ICAe};
\draw (-0.05, 5) -- (0.05, 5) node[anchor=east] {\AntEMax};

% vertical axis - end
\draw[->] (9.8,0) -- (9.8,4) node[anchor=east] {};
\draw (9.8, 0.8) node[anchor=west] {\SLAe};

@@

\caption{\textbf{Qualitative model of the dynamics of anthropomorphism}. We
    distinguish three main phases: \emph{initialization}, \emph{familiarization}
    and \emph{stabilization}, preceded by a \emph{pre-interaction} phase. The
    pre-interaction phase is characterized through an \emph{initial capital of
    anthropomorphism} (ICA, noted \ICA).  Once the interaction starts, the level
    of anthropomorphism increases due to the \emph{novelty effect}, and then
    decreases to reach a \emph{stabilized level of anthropomorphism} (SLA, noted
    \SLA). \ICA and \SLA may vary depending on the user (and his/her
    expectations), the robot and the context of interaction.  During the
    interaction, unpredicted behaviors of the robot (\emph{disruptive
    behaviors}) may lead to local increases or decreases of the level of
    anthropomorphic effects. The duration of these phases may vary and the
    overall shape of the curve is unique to each human interacting with a
    specific robot in a given context, as depicted by the alternative curves.}

@@

\subsubsection{Pre-Interaction and Initialization Phase\\}

The \emph{initial capital of anthropomorphism} \ICA describes the initial potential
for the human user to anthropomorphize the robot in a given situation.
This potential depends on several factors, introduced in
section~\ref{sec:multi-factors}.
It represents the level of anthropomorphic effects that can be expected to occur
during the very first contacts between the human and the robot. During this first
phase, pre-conceptions and initial expectations as well as previous experience
of the user with other similar types of robots and similar contexts are the
primary source of influence.

During the initialization phase, we expect an increasing value of \Ant
that is due to the so-called \textit{effects of novelty} of the
robot. Such effects have been described with interactive technologies in general
\citep{rogers_diffusion_1995}, and also in several long-term HRI studies in
workplaces \citep{huttenrauch_fetch-and-carry_2003,mutlu_robots_2008}, in
schools and public places
\citep{gockley_designing_2005,kanda_communication_2005,kanda_interactive_2004},
and elder care centers \citep{sabelli_conversational_2011}. Also long-term
studies of domestic robots have shown similar effects and found that after some
time people's interest, engagement, and fascination of the robot decreased
\citep{sung_robots_2009,sung_domestic_2010,fernaeus_how_2010,fink_living_2013}.

The type of anthropomorphic effects typically observed during the initialization
hpase include users who spontaneously tend to talk directly to the robot while
they are aware that the system is not be able to recognize speech: greeting the
robot, asking questions to it, giving it a nickname, praising the robot,
commenting on its performances and giving commands. Also, the use of pointing
gestures and waving at the robot are often found, as well as ascribing gender
and some personal preferences or intention to the robot.

@@

In this context, \emph{familiarization} means \emph{to get acquainted with}.
It is not to be confused with the \emph{familiarity thesis} that relates to the
projection of already known cognitive models onto the robot. We discuss this other
aspect in section~\ref{sec:cognition-neuroscience}.

The \emph{familiarization phase} starts at the peak of the \emph{novelty
effect}: anthropomorphic effects are at their maximum, the user thinks (s)he is
potentially facing an agent aiming at ``human-level intelligence'' (to take
McCarthy's words). This is a transient state that quickly vanishes, and the
projected anthropomorphism then starts to decrease while the human observes and
recognizes that the behavior of the robot is generally \emph{predictable} and
possibly \emph{deceptive} (for instance, it talks but does not understand when
we talk; it has eyes but does not recognize everyday objects we show; \etc).

@@

We call \emph{disruptive behaviors} any behavior exhibited by the robot that
appears unexpected to the user: for instance, a robot may usually follow always
the same route to go from one place in a house to the other, but suddenly it
might change the route. The actual underlying reason may span from a bug to the
detection of a new obstacle, but as long as this reason is not immediately
intelligible to the user, the behavior counts as \emph{disruptive}. We further
discuss the role of disruptive behaviors in cognitive terms in section
\ref{sec:disruptive}.

@@

During the \textit{Stabilization Phase} the value of \Ant tends to stabilize over
a longer time, to reach a value that we call the \textit{Stabilized Level of
Anthropomorphism} ($SLA$), noted \SLA. The value that the
$SLA$ approaches is unique for each user interacting with a specific robot in a
given situation. \SLA may approach $0$, when no anthropomorphic effects
are observed anymore, for instance when the user is interacting with the robot
in a routine way and he/she does not project any human-like traits unto the
robot, what \cite{takayama_toward_2011} calls the robot being
\textit{``invisible in use''}. \SLA may as well remains above 0 if the user
continues to exhibit anthropomorphic behaviors when interacting with the robot
on the long run.

The level \SLA, like \ICA, is the result of a combination of multiple human,
robotic and contextual factors, with the additional impact of the
\emph{interaction history t} that the user experienced with the robot. Note that
the $ICA$ and $SLA$ levels are generally not correlated: a person may initially
have a high potential of anthropomorphizing (high \ICA), for instance a highly
anthropomorphic robot, but subsequently get disappointed by the actual
capabilities of it and stop explaining it in terms of human-like characteristics
(low \SLA). Another user may however create lasting affective bonds with the
same robot and may continue to anthropomorphize the robot (higher \SLA).

@@

%\path[fill=gray!20] (0,0) rectangle (10.5,2);
%\path[fill=gray!50] (0,2) rectangle (10.5,4);
%\path[fill=gray!20] (0,4) rectangle (10.5,6);

\draw (-0.3,2) node[anchor=east] {Stage 1}
      (-0.3,4) node[anchor=east] {Stage 2}
      (-0.3,6) node[anchor=east] {Stage 3};

\draw[thin,black!50,dotted] (-0.10,2) -- (1,2);
\draw[thin,black!50,dotted] (-0.10,4) -- (4,4);
\draw[thin,black!50,dotted] (-0.10,6) -- (7,6);

@@

\draw (11,2) node[align=left, anchor=west]{\scriptsize pre-cognitive\\\scriptsize anthropomorphism}; %label

\draw (11,4) node[align=left, anchor=west] {\scriptsize{projection of existing}\\\scriptsize{mental models}\\\scriptsize{\it (familiarity)}}; %label

\draw (11,6) node[align=left, anchor=west] {\scriptsize{adapted mental
model} \\ $\to$ \scriptsize{adapted interaction}\\\scriptsize{\it (anthropomorphic or not)}}; %label

@@

\end{tikzpicture}
%}
\caption{\textbf{The three cognitive stages of anthropomorphism}: Stage 1 is the instinctive,
sub-cognitive identification of living peers. {\it Empathy} is characteristic
of this stage. After longer observation or short, non-contextualized interaction
(typically, a lab environment), the user reaches Stage 2: the user projects a
mental model he/she is already familiar with onto the robot. After longer {\it
contextualized} interaction (typically, at home), the user enters Stage 3 of
anthropomorphism: the user composes a custom mental model of the robot,
based on experience. This leads to adapted interaction modalities, that may
still be anthropomorphic, or not. Note that the different stages coexist: Stage
1 and 2 still affect the human perception of the robot in a Stage 3 interaction.}
\label{fig:cognitivemodel}
\end{figure}

We propose three different cognitive stages (Figure~\ref{fig:cognitivemodel}),
which are not mutually exclusive but overlap over time. Note that these three
cognitive stages are related but do not exactly match the \emph{Initialization},
\emph{Familiarization} and \emph{Stabilization} phases introduced in our model
of the dynamics of anthropomorphism. We discuss this relation hereafter.

@@

The intuitive anthropomorphization in the cognitive Stage 1 might be
characteristic of what \cite{takayama_perspectives_2012} calls the
\textit{in-the-moment} perspective on agency with a robot compared to the
\textit{reflective} perspective (see paragraph ``artifact-centered perspective''
in section \ref{sec:anthropomorphism}).  For instance, we reflectively might
\textit{not} perceive agency in a social robot, but it can feel quite
differently in the moment of interaction. \cite{takayama_perspectives_2012}
argues that neglecting to separate reflective perspectives from in-the-moment
perspectives of agency is one of the major sources of confusion when people talk
and write about anthropomorphism. There seems to be a disconnection between what
people consciously perceive and how they respond to stimuli that they may not
consciously perceive. This means that in the initial phase of interacting with a
novel device \textit{(initialization)} (see Figure \ref{fig:dynamics}) people
might not respond consciously but rather mindlessly \citep{nass_machines_2000},
and in turn anthropomorphize more. They start to respond in a more reflective
manner only during the \textit{familiarization} (in turn, the tendency to
anthropomorphize might decrease). This can be illustrated by the fact that
people tend to deny interacting with computational systems as if they were
people and yet they respond to computers in many ways that are remarkably
similar to how they respond to people \citep{reeves_media_1996}.
\cite{takayama_perspectives_2012} also applies a cognitive viewpoint on the two
different perspectives of in-the-moment \vs reflective to illustrate
differences. She states that in-the-moment perceptions of agency are largely
shaped by bottom-up perceptual processes, evoking very immediate responses (\eg
to the cues emitted by the artifact's design (``artifact-centered
perspective''). In contrast, according to Takayama, reflective perceptions are
more often shaped by top-down processes because of the nature of reflective
thought.

After a longer observation period (typically including complete action sequences
of the robot) or short interaction (touching, short talk like greetings), we
suggest the human reaches the cognitive \emph{Stage 2}: in this phase, the human
starts building a behavioral and cognitive model of the robot that would support
both the observed and imagined capabilities of the robot.  The \emph{familiarity
thesis}~\citep{hegel_understanding_2008} supports the idea that the human first
projects onto the robot mental models of similar agents he/she is already
familiar with (ranging from animals to human adults, to pets and children). We
hypothesize that the nature of the projected mental model, as well as how deep
the human engages in this projection, might be driven by the same parameters as
we mentioned for the \emph{initial capital of anthropomorphism}. We also believe
that, at this stage, the human also changes his attitude towards the robot, by
focusing his attention to social cue more than to low-level behaviors, which
may in turn reinforce the resonance mechanisms.

@@

This mental process depends on the human understanding of the robot's inner
working, as well as his/her own tendency to anthropomorphize (the
\emph{personality} in ICA factor). At this stage, the \emph{perception} of the
robot (its shape for instance) and its original intended \emph{purpose} play
less of an important role. It is mostly a human-centric process.  The result of
this third phase is an iteratively adapted cognitive model of the robot.

Another way to think of the three proposed cognitive stages is to relate them to
the \textbf{user experience} which is characteristic of these three stages. For
instance, \cite{norman_emotional_2003} characterizes  three user experience
dimensions (related to emotional attachment), which occur when people use or see
a product for the first time: \emph{(1)} the \emph{visceral level}, which is the
first impression of a product based on its experience; at this level people do
not think about a product but make spontaneous judgments; \emph{(2)} the
\emph{behavioral level}, in which people use and experience the product,
appraise its functions, and consider aspects such as usefulness and usability;
\emph{(3)} the \emph{reflective level}, in which consciousness takes part in the
process, and past experiences are taken into account.

@@

\begin{figure}[]
\centering
%\resizebox{\linewidth}{!}{
\begin{tikzpicture}
\baselineskip=8pt

\draw[dotted, black!50, thin] (1.8, 6) -- (1.8, 0);
\draw[dotted, black!50, thin] (6.5, 6) -- (6.5, 0);

\draw(0.5,5.5) node[anchor=south west] {\scriptsize \sc Init.};
\draw(2.5,5.5) node[anchor=south west] {\scriptsize \sc Familiarization};
\draw(7,5.5) node[anchor=south west] {\scriptsize \sc Stabilization};

\begin{scope}[yscale=-1,shift={(-0.125,-0.46)}]
    \draw[ultra thick, black!50] svg[scale=1cm] "M 0.125,-1.622832 c 0.093536,-0.9062795
    0.21996726,-1.9131565 0.9017912,-2.5842389 0.4277536,-0.3944265
    1.1653486,-0.5108382 1.605386,-0.071103 1.1558512,0.9403904
    1.9535208,2.2825896 3.2616057,3.0430907 0.871821,0.52801518
    1.8461798,0.91299355 2.8702084,0.98695145 0.4522402,0.0444213
    0.9069132,0.0552179 1.3610087,0.0620963";

\draw (-0.3,1.5) node[anchor=east] {Stage 1}
      (-0.3,3) node[anchor=east] {Stage 2}
      (-0.3,4.5) node[anchor=east] {Stage 3};

\draw[thin,black!50,dotted] (-0.10,1.5) -- (1,1.5);
\draw[thin,black!50,dotted] (-0.10,3) -- (4,3);
\draw[thin,black!50,dotted] (-0.10,4.5) -- (7,4.5);

\draw[->] (0,0) -- (11,0) node[anchor=north] {$t$};
\draw[-|] (0,0) -- (0,6.5) node[anchor=east] {};
% Us
\draw[ultra thick, ->] (0,0) -- (0.2,1.5) -- (1,1.5) -- (1.2,3) -- (3,3) -- (6,4.5) --
(9.5,4.5);
\draw[ultra thick, dashed, ->] (1,1.5) -- (9.5,1.5);
\draw[ultra thick, dashed, ->] (3,3) -- (9.5,3);

\end{tikzpicture}
%}
\caption{\textbf{Towards a cognitive interpretation of the dynamics of
    anthropomorphism}: This diagram combines the figures~\ref{fig:dynamics}
    and~\ref{fig:cognitivemodel} to evidence their interplay: we suggest that
    the cognitive stages 1 and 2 (pre-cognitive anthropomorphism and projection
    of known existing cognitive models) first play a key role in the early
    perception of the robot, leading to high levels of anthropomorphic
    behaviors. Then, the user smoothly transitions from projecting already known
    models to building a new, specific cognitive model of the robot: this occurs
    while the user familiarizes with the robot and leads over time to a
    stabilized level of anthropomorphic behaviors. In the current state of
    robotic technology, this stabilization mostly reads as a \emph{decrease} of
    anthropomorphic perception, due to the user being able to explain and
    accurately predict the robot behavior after enough interaction time, hence
    perceiving the robot as a simple machine.}

\label{fig:mix}
\end{figure}
